(1) Local adaptation and dispersal evolution interact to drive population response to climate change \\ (2) Spatial population structure drives extinction dynamics in climate-induced range shifts \\ (3) Initial spatial structure drives population response to climate change

### 2018-07-23 Manuscript/Main.tex
Updating the final graphing scripts and creating a manuscript draft for Ecology Reports

\documentclass[11pt, oneside]{article}
\usepackage{geometry}                		
\geometry{letterpaper}                   		
\usepackage{graphicx}
\usepackage{graphics}
\usepackage{hyperref}
\usepackage{fancybox}
\usepackage[centertags]{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}			
\usepackage{natbib}				
\usepackage{fullpage}
\usepackage{placeins}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{color}

%\usepackage{figcaps}
%\usepackage[tablesfirst,nolists]{endfloat}
\usepackage{authblk}

@@

\title{Local adaptation and dispersal evolution interact to drive population response to climate change}

@@

\author[1]{Christopher Weiss-Lehman}
\author[1]{Allison K. Shaw}

\affil[1]{Ecology, Evolution, and Behavior, University of Minnesota}

@@

@@

\section{Introduction}
Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global, penuelas2003global, hansen2001global, scholze2006climate}. These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}. In fact, contemporary range shifts have already been observed in a wide range of taxa~\citep{chen2011rapid, walther2002ecological, parmesan2003globally, parmesan2006ecological, parmesan1999poleward, perry2005climate}. Such range shifts present significant challenges to current and future conservation efforts as they can result in both the extinction of populations failing to track a changing climate~\citep{davis2001range, parmesan2006ecological, zhu2012failure, sekercioglu2008climate}, and the creation of novel species assemblages as not all species shift their ranges at the same rate or at all~\citep{hobbs2009novel, gilman2010framework, williams2007novel}, or both. It is therefore crucial to understand the dynamics of such climate-induced range shifts to better inform current and future conservation work.

While the study of range shifts due to climate change is relatively new, important insights can be gained from the related but distinct process of range expansion. Range expansions have been studied for decades, leading to a robust understanding of both the ecological~\citep{skellam1951random, hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, burton2010trade, excoffier2009genetic, kubisch2014and} mechanisms responsible for shaping such expansions. For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, phillips2015evolutionary, shaw2015dispersal}. Since range shifts, like range expansions, have a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions~\citep{hargreaves2014evolution}. However, while such insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have significant differences as well.

In particular, range expansions and range shifts due to climate change involve fundamentally distinct initial conditions. Range expansions, especially for invasive and reintroduced species, typically begin from the successful establishment and subsequent spread of a small, founding population~\citep{hastings2005spatial}. Such founding populations often represent samples from some larger source population and as such lack any initial spatial population structure. In contrast, range shifts start from entire populations with existing spatial structure within the initially stable range~\citep{hargreaves2014evolution}. Such spatial structure can take the form of local adaptation within the range, the severity of the gradient in population size from the range core to edge, or some combination of the two~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.

These different aspects of spatial population structure have been shown to dramatically impact the response of populations to climate change. For example, the mechanism responsible for the gradient in population size from core to edge (e.g. declines in birth rates vs. increases in extinction risk) can directly change the probability of extinction a species faces during a climate driven range shift~\citep{henry2013eco}. A population's risk of extinction during a range shift has also been related to local adaptation within the range. Specifically, a broad environmental niche (i.e. little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted phenotypes~\citep{atkins2010local}. Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal probability in an asexual species as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}. However, it is unclear how dispersal evolution and local adaptation might interact in a sexually reproducing species in which dispersal and fitness are directly linked via gene flow. For example, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range. In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}. Additionally, the distribution of dispersal phenotypes within a population can be influenced by the severity of the gradient at the range edge~\citep{henry2013eco, hargreaves2014evolution}, further complicating the relationship between dispersal evolution and local adaptation during climate-induced range shifts. Evolution during range shifts due to local adaptation and dispersal evolution are both potentially important drivers of range shift dynamics~\citep{van2016spatial} and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.

Here, we develop a complex, individual-based model of a sexually reproducing species, which incorporates evolution in both fitness (via local adaptation) and dispersal distance to determine the interaction between the strength of local selection and the starkness of the gradient at the range edge in driving population dynamics during climate change. Using this model, we simultaneously vary the strength of local selection and the starkness of the range boundary to ascertain the relationship between them and test how dispersal evolution interacts with local adaptation to help or hinder a population's ability to track a changing climate. We additionally compare the dynamics of successful versus doomed simulated populations to understand the factors contributing to population extinction during climate-induced range shifts.

\section{Methods}
\subsection{Overview}
\subsubsection{Purpose}
This model tests an evolving population's ability to track a changing climate under a variety of conditions. Specifically, populations are simulated under different combinations of (1) the starkness of the range boundary and (2) the strength of local selection. In all simulations, individual dispersal and relative fitness are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time. All simulations begin with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change. Climate change is then modeled as a constant, directional shift in the location of environmentally suitable habitat (see the \textit{Submodels} section below). Finally, simulations end with another short period of climate stability to assess the population's ability to persist and recover after shifting its range.

\subsubsection{State variables and scales}
The model simulates a population of male and female individuals characterized by spatial coordinates for their location and by diploid loci for both fitness and dispersal. Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system with a fixed width along the y axis and without bounds on the x axis. To avoid edge effects along the y axis, the model employs wrapping boundaries such that if an individual disperses out of the landscape on one side, it appears at the opposite end of the same row of the landscape. Patches are defined by the location of the patch center in x and y coordinates and a patch width parameter defining the relationship between continuous Cartesian space and the discrete patches used for population dynamics (see the \textit{Submodels} section below).

The abiotic environment is defined by the location of the center of the range, the slope of the range boundary, and the width of the range (See Figure~\ref{fig:EnvFunction}). The range can then shift in space by altering the location of the range center, which is how climate change is implemented in the model. Further, a gradient of phenotypic optimum values for fitness is imposed within the range to allow for local adaptation within the range boundaries. The severity of this gradient is defined independently of the other environmental parameters.

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
\vspace{-5mm}
\caption[LoF entry]{Environmental quality, as defined by $f(x,t)$, over a spatial gradient. The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis. Additionally, the lattice of discrete patches in which population dynamics occur is shown beneath. This lattice of $\eta$ x $\eta$ patches are mapped to environmental conditions along the $x$ axis using $f(x,t)$ and are environmentally constant along the $y$ axis. Dispersal is unbounded in the $x$ direction and implemented with wrapping boundaries in the $y$ direction.}
\label{fig:EnvFunction}
\end{figure}

\subsubsection{Process overview and scheduling}
Time is also modeled in discrete intervals defining single generations of the population. Within each generation, individuals first disperse from their natal patches according to their dispersal phenotypes (\textit{Submodels}). After dispersal, reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch (\textit{Submodels}). Parental pairs form via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with high fitness are likely to produce multiple offspring while individuals with no fitness may not produce any. Individuals inherit one allele from each parent at each loci, assuming independent segregation and a mutation process (\textit{Submodels}). After reproduction, all individuals in the current generation perish and the offspring begin the next generation with dispersal, resulting in discrete, non-overlapping generations.

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
\vspace{-5mm}
\caption[LoF entry]{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal. Each generation begins with new offspring dispersing according to their phenotype, after which reproduction occurs in local populations defined by the discrete lattice. After reproduction, all parental individuals perish, resulting in discrete, non-overlapping generations.}
\label{fig:LifeCycle}
\end{figure}

\subsection{Design concepts}
\subsubsection{Emergence}
Emergent phenomena in this model include the spatial equilibrium of population abundances and trait values within the stable range, the demographic dynamics of the shifting population during climate change, and the evolutionary trajectories of both fitness and dispersal traits during climate change. These are all examined in the context of their impact on the population's ability to keep pace with the changing climate.

\subsubsection{Stochasticity}
All biological processes in this model are stochastic including realized population growth in each patch, dispersal distances of each individual, and inheritance of loci. Environmental parameters are fixed, however, and the process of climate change (i.e. the movement of environmentally suitable habitat through time) is deterministic. Thus, the model removes the confounding influence of environmental stochasticity to focus on demographic and evolutionary dynamics of range shifts.

\subsubsection{Interactions}
Individuals in the model interact via mating and density-dependent competition within patches. Other important interactions are the relationship between dispersal evolution and local adaption, particularly in edge populations, and how this relationship impacts a population's ability to avoid extinction and track a changing climate. These interactions are examined in the context of different configurations of the initial, stable range and different speeds of climate change.

\subsubsection{Desired output}
After each model run, full details of all surviving individuals at the last time point are recorded (spatial coordinates and loci values for both traits). If a population went extinct during the model run, the time of extinction is recorded. Throughout the simulation, certain aggregated values are calculated and recorded for each occupied patch including the population size and the mean and variance of allele values for each trait.

\subsection{Details}
\subsubsection{Initialization}
The following parameters are set at the beginning of each simulation and form the initial conditions of the model: the mean and variance for allele values of each trait, population size, location of environmentally suitable habitat, number of generations for the pre-, post-, and during climate change periods of the simulation, and all other necessary parameters for the submodels defined below. Simulated populations are initialized in the center of the range and allowed to spread and equilibrate throughout the range during the period of stable climate conditions. This ensures that the populations reacting to a changing climate truly represent the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation. See Table 1 for a full list of all parameter values used in the simulations described here.

\subsubsection{Submodels}
\paragraph{Environmentally suitable habitat}
Environmentally suitable habitat is determined by the population's carrying capacity as it ranges in space ($K_{x}$). The carrying capacity is maximized in the center of the species' range ($K_{max}$) and declines with increasing distance from the center. Specifically, the carrying capacity at a location $x$ is defined as the product of $K_{max}$ and a function $f(x,t)$, where $f(x,t)$ ranges from $1$ in the range center to $0$ far away from the center and is defined as follows 
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau)}} & x \leq \beta_{t} \\
	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau)}} & x > \beta_{t}
\end{cases}
\end{equation}
where $\beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure~\ref{fig:EnvFunction}). To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the spatial dimension can be shown to be
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{\gamma e^{\gamma(x-\beta_{t}+\tau)}}{(1+e^{\gamma(x-\beta_{t}+\tau})^{2}} & x \leq \beta_{t} \\
	\frac{-\gamma e^{-\gamma(x-\beta_{t}-\tau)}}{(1+e^{-\gamma(x-\beta_{t}-\tau})^{2}} & x > \beta_{t}
\end{cases}	
\end{equation}
yielding a derivative of $\pm\frac{\gamma}{4}$ at the inflection points on either side of the range center ($x=\beta_{t}\pm\tau$).

Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure~\ref{fig:EnvFunction}). The local carrying capacity of a patch centered on $x$ ($K_{x}$) is then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
\begin{equation}
K_{x} = \frac{K_{max}}{\eta}\int_{x-\frac{\eta}{2}}^{x+\frac{\eta}{2}}f(x,t)dx
\end{equation}

By varying the parameters defining $f(x,t)$, we can change both the total achievable carrying capacity of the population throughout the range (by altering both $\tau$ and $\gamma$) and the slope at which $K_{x}$ declines to $0$ (by altering $\gamma$). Changing the slope affects not only the rate at which $K_{x}$ declines at the range boundaries (our focus), but it also alters the total achievable carrying capacity of the population. To avoid this confounding factor, we fix the total area under the curve $f(x,t)$. The indefinite integral of $f(x,t)$ can be shown to be
\begin{equation}
\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1)}{\gamma}
\end{equation}
which can be solved for $\tau$. For a given fixed total area under the curve, an appropriate value of $\tau$ can be calculated for each value of $\gamma$.

Thus, $\gamma$ and $\tau$ are both fixed within a given simulation and $\beta_{t}$ (the location of the center of suitable habitat) is used to simulate climate change. During the periods before and after climate change $\beta_{t}$ is constant, but to simulate climate change it varies with time as follows
\begin{equation}
\beta_{t}=\nu\eta(t-\hat{t})
\end{equation}
where $\nu$ is the velocity of climate change per generation in terms of discrete patches, $\eta$ is the spatial scale of each patch, $t$ is the current generation, and $\hat{t}$ is the last generation of stable climatic conditions before the onset of climate change.

\paragraph{Local adaptation}
To allow an arbitrary degree of local adaptation within the range, the local phenotypic optima ($z_{opt,x}$) is set as follows
\begin{equation}
z_{opt,x}=\lambda(x-\beta_{t})
\end{equation}
where $\lambda$ defines the strength of local selection with values close to $0$ resulting in little to no change in phenotypic optimum across the range and values of greater magnitude resulting in large differences in phenotypic optima across the range. Individual fitness ($w_{i,x}$) values are then calculated according to the following equation assuming stabilizing selection
\begin{equation}
w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
\end{equation}
where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's fitness phenotype~\citep{lande1976natural}. All loci are assumed to contribute additively to the phenotype with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values for the fitness trait.

\paragraph{Population dynamics}
Population growth within each patch is modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}. To account for fitness effects on population growth, expected population growth is scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$). The expected number of new offspring in patch $x$ at time $t+1$ is then given by
\begin{equation}
\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
\end{equation}
where $F_{t,x}$ is the number of females in patch $x$ at time $t$, $R$ is the intrinsic growth rate for the population, $\psi$ is the expected sex ratio of the population, $N_{t,x}$ is the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ is the local carrying capacity based on the environmental conditions. To incorporate demographic stochasticity, the realized number of offspring for each patch is then drawn from a Poisson distribution.
\begin{equation}
N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
\end{equation}

Parentage of the offspring is then assigned by random sampling of the local male and female population (i.e. polygynandrous mating). The sampling is weighted by individual fitness and occurs with replacement so highly fit individuals are likely to have multiple offspring while low fitness individuals may not have any. Each offspring inherits one allele per locus from each parent, assuming no linkage among loci. After reproduction, all members of the previous generation die and the offspring disperse to begin the next generation.

>>>>

\documentclass[12pt, oneside]{article}
\usepackage{geometry}                		
\geometry{letterpaper}                   		
\usepackage{graphicx}
\usepackage{graphics}
\usepackage{hyperref}
\usepackage{fancybox}
\usepackage[centertags]{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{natbib}				
\usepackage{fullpage}
\usepackage{placeins}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{color}
\usepackage{mathptmx}
\usepackage{authblk}

@@

\title{(1) Local adaptation and dispersal evolution interact to drive population response to climate change \\ (2) Spatial population structure drives extinction dynamics in climate-induced range shifts \\ (3) Initial spatial structure drives population response to climate change}

@@

\author[1,*]{Christopher Weiss-Lehman}
\author[1]{Allison K. Shaw}

\affil[1]{Ecology, Evolution, and Behavior, University of Minnesota}
\affil[*]{Denotes corresponding author}

@@

\begin{flushleft}
Corresponding author details: \\
Christopher Weiss-Lehman \\
email: cweissle@umn.edu \\
phone: +1 (720) 590 2278 \\
\end{flushleft}

@@

\newpage

\section*{Abstract}
Anthropogenic climate change has already caused some species' ranges to shift upwards in latitude and elevation, making it imperative to understand the ecological and evolutionary dynamics behind such range shifts. While insights can be gained from the related process of range expansion, climate-induced range shifts have several key attributes differentiating them from range expansions. In particular, range shifting populations are characterized by some initial spatial population structure defined by factors such as local adaptation and the nature of the range edge. Here, we use an individual-based model to explore the importance of spatial population structure for range shifting populations. We show that populations with more pronounced spatial structure have substantially increased extinction probabilities. Further, in our simulations evolution of dispersal ability was insufficient to rescue faltering populations. Rather, a population's fate during climate change was determined by the evolved composition of dispersal phenotypes within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived. Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that the initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.

\begin{flushleft}
\textbf{Key words:} range shifts, eco-evolutionary dynamics, local adaptation, individual-based model
\end{flushleft}

\newpage

\section*{Introduction}
Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global}. These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}. In fact, contemporary range shifts have already been observed in a wide range of taxa~\citep{chen2011rapid, parmesan2006ecological}. Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} the creation of novel species assemblages as not all species shift their ranges at the same rate or at all~\citep{hobbs2009novel}, or both. It is therefore crucial to understand the dynamics of such climate-induced range shifts to better inform current and future conservation work.

While the study of range shifts due to climate change is relatively new, important insights can be gained from the related but distinct process of range expansion. Range expansions have been studied for decades, leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms responsible for shaping such expansions. For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal}. Since range shifts, like range expansions, have a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions~\citep{hargreaves2014evolution}. However, while such insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have significant differences as well.

In particular, range expansions and range shifts due to climate change involve fundamentally distinct initial conditions. Range expansions, especially for invasive and reintroduced species, typically begin from the successful establishment and subsequent spread of a small, founding population~\citep{hastings2005spatial}. Such founding populations often represent samples from some larger source population and as such lack any initial spatial population structure. In contrast, range shifts start from entire populations with existing spatial structure within the initially stable range~\citep{hargreaves2014evolution}. Such spatial structure can manifest through local adaptation within the range, the severity of the gradient in population size from the range core to edge, or some combination of the two~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.

These different aspects of spatial population structure have been shown to dramatically impact the response of populations to climate change. For example, the mechanism responsible for the gradient in population size from core to edge (e.g. declines in per capita growth rate vs. increases in local extinction probability) can directly change the probability of extinction a species faces during a climate driven range shift~\citep{henry2013eco}. A population's risk of extinction during a range shift has also been related to the degree of local adaptation within the range. Specifically, a broad environmental niche (i.e. little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted genotypes~\citep{atkins2010local}. Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal probability in an asexual species as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}. However, it is unclear how dispersal evolution and local adaptation might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow. For example, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range. In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}. Additionally, the distribution of dispersal phenotypes within a population can be influenced by the severity of the gradient at the range edge~\citep{henry2013eco, hargreaves2014evolution}, further complicating the relationship between dispersal evolution and local adaptation during climate-induced range shifts. Evolution during range shifts due to local adaptation and dispersal evolution are each potentially important drivers of range shift dynamics and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.

Here, we develop a two-sex individual-based model with two genetically determined traits, one defining an individual's expected dispersal distance and the other determining an individual's environmental niche. Using this model, we vary the degree of spatial population structure defining the stable range to ascertain how this impacts a population's ability to track a changing climate. Further, we contrast the dynamics of extant and extinct populations to understand the joint impact of dispersal evolution and local adaptation on extinction risk during climate-induced range shifts.

\section*{Methods}
A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in the online supplemental materials, but we present a brief summary here. Each simulated landscapes consisted of a discrete lattice of habitat patches with a fixed width in the $y$ dimension and unbounded length in the $x$ dimension. Environmental conditions varied along the $x$ dimension and remained constant across the $y$ dimension (Fig. S1). To simulate climate change, environmental conditions shifted a constant amount in the $x$ dimension each generation. Competition and reproduction occurred within each patch and local populations were linked via dispersal, assuming wrapping boundaries in the $y$ dimension to prevent edge effects.

Individuals were characterized by two traits (dispersal and an environmental niche), both defined by a set of $5$ quantitative diploid loci, assuming no linkage. The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel. An individual's environmental niche trait allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's location, the higher the individual's realized fitness. The potential for local adaptation was then manipulated by varying the degree to which the environmental optimum changed throughout the landscape. In addition to the potential for local adaptation, simulated ranges were characterized by a decline in patch carrying capacity from the range center to the edge, the severity of which could be adjusted without altering the total carrying capacity of the landscape (see supplemental materials; Fig. S1). Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}. Using this model, we simulated non-overlapping generations consisting of discrete dispersal and reproduction phases (Fig. S2) to determine the effect of different aspects of spatial population structure on the eco-evolutionary dynamics of climate-induced range shifts.

Most model parameters were held constant over all simulations (Table S1), but we varied certain parameter values to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table S2). Specifically, we considered a factorial combination of no, low, and high potential for local adaptation with shallow, moderate, and stark gradients in suitable habitat at the range edge for a total of $9$ different scenarios, each explored with $200$ simulations. Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions. Figure~\ref{fig:SimExample} shows an example of a single simulation undergoing climate change. Additionally, we explored the role of the rate of climate change on the dynamics of range shifts in these different scenarios, examining both slower and faster speeds of climate change (see the online supplemental materials).

For each parameter combination, we calculated the per patch change in mean dispersal phenotype from the beginning of the period of climate change to the end to quantify dispersal evolution. For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates. In this way, we calculated change in mean dispersal phenotype for populations consistently occupying the same position relative to the range center (e.g. leading edge vs. core populations). Due to local extinctions, not all patches were occupied at the end of the period of climate change. To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals as the end point to calculate change in the mean dispersal phenotype. Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions) so that positive values indicate an increase in the mean dispersal phenotype. All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at \url{https://github.com/tpweiss06/ShiftingSlopes}.

\section*{Results}
For almost all parameter combinations examined here, a portion of the simulated populations were able to keep pace with changing climate conditions. Successful populations shifted their spatial distributions at essentially the same rate as climate change (e.g. Fig. 1). However, some populations failed to keep pace, lagging further behind the shifting climate until they went extinct. Populations defined by a higher potential for local adaptation and by stark habitat gradients at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig. 2). While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction dynamics. The parameters defining both of these range attributes were varied widely (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table S2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space. Additionally, the pace of climate change also influenced extinction probabilities as expected with faster climate change corresponding to greater extinction risk (Fig. S3 \& S4). However, this effect was independent of the roles of local adaptation and the habitat gradient at the range edge in determining the extinction probability for a range shifting population.

Dispersal evolution is expected to aid populations shifting spatially to track a changing climate, and, as expected, the average dispersal phenotypes increased through time in some simulations (e.g. Fig.~\ref{fig:SimExample}). However, comparing patterns in dispersal evolution between simulated populations that successfully tracked climate change and those that ultimately went extinct revealed no differences in the magnitude or direction of dispersal evolution (Fig. 3a\&b). Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes to dispersal phenotypes strongly centered on $0$ (Fig. S5-S7). The similarity in evolved changes to dispersal between surviving and extinct simulated populations suggests that this cannot explain the different outcomes achieved by these populations.

Instead, the initial distribution of dispersal phenotypes immediately before the beginning of climate change played a large role in determining the ultimate fate of simulated populations. Simulated populations evolved a range of dispersal phenotypes over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig. S8-S10). The populations that survived climate change were those that happened to make up the upper end of these distributions, containing individuals with high dispersal phenotypes (Fig. 3c\&d). In fact, comparing dispersal phenotypes of successful populations across different parameter combinations revealed a threshold in dispersal phenotypes separating individuals from successful and extinct populations (Fig. S8-S10). This threshold varies directly with the speed of climate change, increasing with more rapid climate change. Across all speeds of climate change, this threshold corresponds to roughly similar probabilities ($12\%$, $9\%$, and $7\%$ at slow, moderate, and fast speeds respectively) of an individual keeping pace with climate change (i.e. moving the correct number of patches in the $x$ dimension to track the changing environmental conditions).

Intriguingly, the simulations that survived climate change tended to be characterized by reduced fitness at the range margins compared to simulated populations that went extinct (Fig. S11-S13). This pattern was most evident in the simulations with either (1) a gradual environmental gradient at the range edge or (2) a high potential for local adaptation. Intuitively, there was no spatial pattern in fitness for simulations with no potential for local adaptation. As the simulations that survived climate change were also characterized by heightened dispersal (Fig. 3), the observed reduction in average patch fitness at the margins is likely due to increased gene flow from the range core hampering the abilities of these edge populations to adapt to local conditions.

\section*{Discussion}
Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts. Here we developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci. In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g. single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}). Thus, in analyzing our model, we tested the generality of previous results under realistic levels of biological complexity. Specifically, we demonstrated the role of spatial population structure, in the form of local adaptation and the environmental gradient defining the range edge, in determining extinction risk for range shifting populations via impacts on the initial distribution of dispersal phenotypes and environmental niche values.

Our results suggest that populations most likely to keep pace with climate change will be those with little to no local adaptation within the range and in locations with shallow environmental gradients defining the range edge (Fig. 2). A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies testing for it, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}. Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the stark environmental gradients imposed in our study. While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, combined with our results they suggest many species will face elevated extinction risks in climate-induced range shifts due to the spatial population structure of their initial ranges.

A key result of our model was the importance of the initial distribution of dispersal phenotypes (i.e. prior to the start of climate change) in determining a population's extinction risk during climate change (Fig. 3c\&d). Survival in the face of climate change was primarily determined by whether or not simulated populations consisted of individuals with dispersal phenotypes at or above some threshold value. Importantly, the threshold necessary to survive climate change itself was constant across all parameter combinations at a given speed of climate change (Fig. S8-S10). The differences in extinction probabilities among different scenarios were instead driven by the overall distribution of dispersal phenotypes that evolved in different simulations under each parameter combination. Scenarios with no potential for local adaptation and gradual environmental gradients simply had larger proportions of high dispersal phenotypes under stable climate conditions. A high potential for local adaptation, in contrast, selected against such high dispersal phenotypes due to dispersal's homogenizing effect on population genetic structure~\citep{lenormand2002gene}. Similarly, a more severe habitat gradient at the range edge increases the risk of dispersing beyond the boundary of suitable habitat, resulting in selection against heightened dispersal~\citep{shaw2014population}. Thus, attributes defining the spatial population structure of simulated ranges resulted in different initial distributions of traits under stable climate conditions, which then determined the extinction risk of populations under climate change.

While high dispersal phenotypes prior to climate change allowed populations to survive, it had the additional effect of reducing average fitness at the range margins in scenarios with the potential for local adaptation (Fig. S11-S13). In the model, populations at the range margins tended to be lower abundance than populations in the range core, thus increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}. Thus, when populations were defined by high dispersal phenotypes prior to climate change, it reduced fitness at the range margin due to gene flow preventing adaptation to local conditions. As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness prior to the start of climate change. While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, and thus are unlikely to be equally susceptible to the influence of gene flow from the core, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.
